---
title: "Funprofiler Data Cleaning"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Set up

```{r}
library(tidyverse)
library(readxl)
library(janitor)
library(gt)
library(ggplot2)
library(viridis)
```

#### Read in files ####
# set directory to folder containing all raw funprofiler output profiles
```{r}
# prevent scientific notation
options(scipen = 999)

# identify sequencing files
file_list <-  list.files(pattern = "profiles")

# create empty df to hold results
df_clean <- data.frame()

# loop through sequencing files

for (file in file_list) {
  df <- read_csv(file) %>%
    mutate(file_id = substr(file, 1, 32)) #This works for S numbers from labvantage, edit for other uses
  df_clean <- rbind(df_clean, df)
  print(file)
}
```


Set the directory to the final repository you want everything stored in
```{r}
df_clean |> 
  write_csv("data/funprofiler/2_clean_profiles/funprofiler_cdi_2.csv.bz2")
#Data is often large, use a .csv.bz2 format to compress size
```

####Clean data for use ####

To create a compiled file with ID's pulled from larger filenames, 
use a list
```{r}
##### Pull out sample ID's and attach metadata #####

#Example file containing sample IDs, this can be pulled from labvantage or 
#running listfiles to gather the names of all samples 
metadata_file<-read_xlsx("data/List files metamira 1-2.xlsx")


metadata_ids<-metadata_file |>
  janitor::clean_names() |>
  select(file_name) |>
  rename(sample_id = current_file_name) |>
  mutate(sample_id = gsub("_R1.fastq.gz", "", sample_id)) |>
  mutate(sample_id = gsub("_R2.fastq.gz", "", sample_id)) |>
  distinct() 

write_csv(metadata_ids, "data/metadata_ids.csv")
```

```{r}
# Create string of sample IDs from existing metadata to pull them out

#List of id's to pull from full filenames
metadata_ids |>
 # mutate(sample_id = gsub("_", "-", sample_id)) |> #if there are different formats between metadata and your output IDs, edit
  pull(sample_id) |>
  paste0(collapse = "|") |>
  identity() -> id_search_string

```

```{r}
df_clean |>
  mutate(sample_id = file_id) |>
  mutate(sample_id = gsub("_","-", sample_id)) |> #Use if formatting is inconsistent 
  mutate(sample_id = str_extract(string = sample_id,id_search_string )) |>
  mutate(read_pair = case_when(str_detect(file_id,"R1") ~ "R1",
                               str_detect(file_id,"R2") ~ "R2")) |> #This is relevant for read-paired data
  left_join(metadata_file, by = "sample_id") |> 
  identity() -> clean_funprofiler

write_csv(clean_funprofiler, filename= "data/clean_funprofiler.csv")
```

##### Average abundances between read pairs #####
Funprofiler runs each read pair individually, so abundance values need to be 
averaged to create one dataset. 
```{r}
clean_funprofiler |>
  group_by(external_participant_id,collection_date,sample_id,sample_type,study,read_pair) |>
  mutate(total_abundance = sum(abundance))|> #Should add up to 1 per read pair
  select(total_abundance, read_pair)

clean_funprofiler |> #Average read pair abundances
  group_by(sample_id, external_participant_id, collection_date, study, sample_type) |> #original_sample_type
  mutate(avg_abundance = mean(abundance)) |>
  select(external_participant_id, collection_date, sample_id, ko_id,avg_abundance, sample_type, study, sample_type) |> 
  distinct() |>
  ungroup() |>
  identity() -> clean_funprofiler_avg_rp

clean_funprofiler_avg_rp |>
  group_by(external_participant_id, collection_date, sample_id,avg_abundance, study, sample_type) |>
  mutate(total_abundance = sum(avg_abundance)) #New total abundances per specimen

clean_funprofiler_avg_rp |> #Convert new abundances into relative abundances
  filter(!is.na(external_participant_id)) |>
  group_by(sample_id, external_participant_id, collection_date, study, sample_type) |>
  mutate(total_sum = sum(avg_abundance)) |>
  ungroup() |>
  group_by(sample_id, external_participant_id, collection_date, study, sample_type, ko_id) |>
  mutate(avg_rel_abundance = (avg_abundance/total_sum)) |>
  ungroup() |>
  identity() -> clean_funprofiler_relative_abs

clean_funprofiler_relative_abs |>
  group_by(sample_id, external_participant_id, collection_date, study, sample_type) |>
  mutate(totals = sum(avg_rel_abundance)) #Check that adds up to 1

write_csv(clean_funprofiler_relative_abs, "data/funprofiler/3_profiles_w_metadata/funprofiler_cdi_2_metadata_clean.csv.bz2")

```

